---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

by Yiannis Manatos, © Feb. 2016


## R libraries initialization

```{r step_init, echo = TRUE}
# R libraries initialization
library(scales)
library(ggplot2)
library(dplyr)

tz <- Sys.getenv("TZ")
Sys.setenv(TZ = "GMT")
```


## I) Loading and preprocessing the data

```{r step_I, echo = TRUE}
# Read data from file
all_data <- read.csv("activity.csv", header = TRUE)
str(all_data)
# Add 'Date' column, from 'date'
all_data$Date <- as.Date(all_data$date, 
                         tz = "GMT")
# Add 'Interval' column, from 'interval'
all_data$Interval <- as.POSIXct(paste("2016-01-01", sprintf("%04d", all_data$interval)), 
                            tz = "GMT", 
                            format = "%Y-%m-%d %H%M")

# Examine all data read
str(all_data)
head(all_data)
summary(all_data)
```


## II) What is mean total number of steps taken per day?

```{r step_II, echo = TRUE}
# Filter out incomplete cases 
good_data <- all_data[complete.cases(all_data),]
str(good_data)

# Calculate the total number of steps taken per day
steps_by_day <- summarise( group_by(good_data, Date), sum(steps) )
str(steps_by_day)
steps_per_day <- steps_by_day[[2]]

# Make a histogram of the total number of steps taken each day
hist( steps_per_day, 
      xlab = "# of steps per day", 
      main = "Total # steps taken each day" )

# Calculate and report the mean and median of the total number of steps taken per day
mean_steps_per_day <- format(mean(steps_per_day))
median_steps_per_day <- format(median(steps_per_day))
```

#### The **mean** of the total number of steps taken per day is: **`r mean_steps_per_day`** .  
#### The **median** of the total number of steps taken per day is: **`r median_steps_per_day`** .


## III) What is the average daily activity pattern?

```{r step_III, echo = TRUE}
# Calculate the total number of steps taken per 5-minute interval (across all days)
steps_by_interval <- summarise( group_by(good_data, Interval), sum(steps), mean(steps) )
str(steps_by_interval)
names(steps_by_interval) <- c("Interval", "steps_sum", "steps_mean")
# Examine the interval-grouped data
str(steps_by_interval)
head(steps_by_interval)
summary(steps_by_interval)

# Make a simple time-series plot of the 5-minute interval (x-axis)
# and the average number of steps taken, averaged across all days (y-axis)
with(steps_by_interval, plot( Interval, steps_mean, type = "l", 
                              xlab = "5-min interval",
                              ylab = "# of steps", 
                              main = c("Average # steps per 5-min interval", 
                                       "across all days", 
                                       "(simple version, see below for a better one)") ))

# Make a better time-series plot, handling intervals scale more accurately
ggplot(steps_by_interval, 
       mapping = aes(Interval, steps_mean)) + 
    geom_line() + 
    labs( x = "5-min interval", 
          y = "# of steps", 
          title = "Average # steps per 5-min interval\nacross all days\n(more accurate intervals version)") +
    scale_x_datetime(breaks = date_breaks("2 hour"), 
                     labels = date_format("%H:%M"))

# Calculate and report the 5-min interval, on average across all the days
# in the dataset, that contains the maximum number of steps
max_i_mean <- max(steps_by_interval$steps_mean)
max_i_row <- steps_by_interval[steps_by_interval$steps_mean == max_i_mean, ]
max_interval <- format(max_i_row[1, 1],
                       tz = "GMT", 
                       format = "%H:%M")
```

#### The **5-min interval** in the dataset containing the **maximum number of steps**, on average across all the days, is: **`r max_interval`** .


## IV) Imputing missing values

```{r step_IV_1, echo = TRUE}
# 1. Calculate and report the total number of missing values in the dataset
# (i.e. the total number of rows with NAs)
na_rows <- sum( is.na(all_data$steps) )
```

#### 1. The **total number of missing values** in the dataset (i.e. the total number of rows with NAs), is: **`r na_rows`** .

```{r step_IV_2, echo = TRUE}
# 2. Devise a strategy for filling in all of the missing values in the dataset.
# The strategy does not need to be sophisticated. For example, you could use the mean/median for
# that day, or the mean for that 5-minute interval, etc na_rows <- sum( is.na(all_data$steps) )
```

#### 2. The strategy chosen to follow for filling in all of the missing values in the dataset is to use **the rounded mean for the 5-minute interval**.

```{r step_IV_3, echo = TRUE}
# 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
new_data <- all_data[, c("steps", "Date", "Interval")]
str(new_data)
# Fill in missing values, using the rounded mean of the 5-minute interval
n <- nrow(new_data)
for (i in 1:n)
    if (is.na(new_data[i, 'steps']))
    {
        t <- new_data[i, 'Interval']
        m <- steps_by_interval[steps_by_interval$Interval == t, 
                               'steps_mean']
        new_data[i, 'steps'] <- round(m)
    }
# Examine all data read
str(new_data)
head(new_data)
summary(new_data)
```

```{r step_IV_4, echo = TRUE}
# 4. Make a histogram of the total number of steps taken each day
new_steps_by_day <- summarise( group_by(new_data, Date), sum(steps) )
str(new_steps_by_day)
new_steps_per_day <- new_steps_by_day[[2]]
hist( new_steps_per_day, 
      xlab = "# of steps per day", 
      main = c("Total # steps taken each day", "(after NAs imputing)") )

# Calculate and report the mean and median total number of steps taken per day.
# Do these values differ from the estimates from the first part of the assignment? 
# What is the impact of imputing missing data on the estimates of the total daily number of steps?
new_mean_steps_per_day <- format(mean(new_steps_per_day))
new_median_steps_per_day <- format(median(new_steps_per_day))
```

#### The new **mean** of the total number of steps taken per day is: **`r new_mean_steps_per_day`** (before NAs imputing it was **`r mean_steps_per_day`**) .  
#### The new **median** of the total number of steps taken per day is: **`r new_median_steps_per_day`** (before NAs imputing it was **`r median_steps_per_day`**) .


## V) Are there differences in activity patterns between weekdays and weekends?

```{r step_V, echo = TRUE}
# 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
# indicating whether a given date is a weekday or weekend day.
#new_data$Weekday <- factor(NA, levels = c("weekday", "weekend"))
#new_data$Weekday <- weekdays(new_data$Date)
weekday <- "weekday"
weekend <- "weekend"
days <- c(weekend, 
          weekday, weekday, weekday, weekday, weekday, 
          weekend)
new_data$Weekday <- as.factor( days[ as.POSIXlt(new_data$Date)$wday + 1 ] )
# Examine new_data updated
str(new_data)
head(new_data)
summary(new_data)

# 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-min interval (x-axis) 
# and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
# See the README file in the GitHub repository to see an example of what this plot should look like.

# Calculate the total number of steps taken per 5-minute interval (across all days)
new_steps_by_interval <- summarise( group_by(new_data, Weekday, Interval), sum(steps), mean(steps) )
str(new_steps_by_interval)
names(new_steps_by_interval) <- c("Weekday", "Interval", "steps_sum", "steps_mean")
# Examine the interval-grouped data
str(new_steps_by_interval)
head(new_steps_by_interval)
summary(new_steps_by_interval)

# Make the 2-facet plot
ggplot(new_steps_by_interval, 
       mapping = aes(Interval, steps_mean)) + 
    geom_line() + 
    facet_grid(Weekday ~ .) +
    labs( x = "5-min interval", 
          y = "# of steps", 
          title = "Average # steps per 5-min interval\nacross all days\n(accurate intervals grouped version)") +
    scale_x_datetime(breaks = date_breaks("2 hour"), 
                     labels = date_format("%H:%M"))
```


## -- END-OF-REPORT --
```{r step_end, echo = TRUE}
Sys.setenv(TZ = tz)
```
